Reduced plasticity and variance in physiological rates of ectotherm populations under climate change

### 2024-02-14 ms.qmd
Abstarct and afflication fixed

---
title: "Reduced plasticity and variance in physiological rates for ectotherm populations under climate change"
bibliography: ./bib/refs.bib
csl: ./bib/nature.csl
format:
  docx:
    reference-doc: ./bib/template.docx
editor_options: 
  chunk_output_type: console
execute:
  freeze: auto  # re-render only when source changes
  cache: false
  echo: false
  warning: false
  error: false
  include: true
crossref:
  fig-title: 'Figure'
  fig-labels: arabic
  title-delim: "-"
  fig-prefix: "Figure"
  tbl-prefix: "Table"
---

@@

#### Affliations:
^1^ Division of Ecology and Evolution, Research School of Biology, The Australian National University, Canberra, ACT 2600, Australia &NewLine;  
^2^ Ecology and Evolution Research Centre, School of Biological, Earth and Environmental Sciences, University of New South Wales, Sydney, NSW, Australia &NewLine;  
^3^ SOLES, University of Sydney, Sydney, NSW, Australia &NewLine;  
^4^ Department of Biology, Lancaster University, Liverpool, UK &NewLine;  
$\dagger$ contributed equally &NewLine;  
$\ddagger$ corresponding author, daniel.noble@anu.edu.au &NewLine;

@@

Climate change is expected to result in warmer and more variable thermal environments globally. Greater thermal variability is expected to result in strong selection pressures leading to genetic adaptation and/or the evolution of adaptive phenotypic plasticity. Such responses depend on genetic and phenotypic variability, yet most work has focused on plastic changes in mean phenotypic responses to climate warming ignore how temperature may change phenotypic variability. Using a large database of physiological rate data from marine, freshwater and terrestrial ectotherms we quantify the capacity for temperature acclimation and the extent to which relative variance (i.e., coefficient of variation) in physiological rates change at high temperatures. Using new effect size estimates and meta-analysis, we show that across habitats relative variance in physiological rates decreased at higher temperatures. Freshwater ectotherms are capable of acclimating and have the smallest reductions in relative variance. Marine organisms also showed a capacity to acclimate to higher temperatures but they coincide with a reduction in relative variance in physiological rates. Relative variance reductions were particularly pronounced for terrestrial ectotherms, and this coincides with a lack of capacity for acclimation highlighting the vulnerability of terrestrial ectotherms to climate change. Neither life-history stage nor past climate explained effect variability. Our results emphasise how beneficial acclimation responses may trade-off with reductions in physiological rate variance that could have important evolutionary and ecological ramifications that affect our understanding of how climate change will impact populations in the future.

>>>>

---
title: "Reduced plasticity and variance in physiological rates of ectotherm populations under climate change"
bibliography: ./bib/refs.bib
csl: ./bib/nature.csl
format:
  docx:
    reference-doc: ./bib/template.docx
editor_options: 
  chunk_output_type: console
execute:
  freeze: auto  # re-render only when source changes
  cache: false
  echo: false
  warning: false
  error: false
  include: true
crossref:
  fig-title: 'Figure'
  fig-labels: arabic
  title-delim: "-"
  fig-prefix: "Figure"
  tbl-prefix: "Table"
---

@@

#### Affliations:
^1^ Division of Ecology and Evolution, Research School of Biology, The Australian National University, Canberra, ACT 2600, Australia &NewLine;  
^2^ Ecology and Evolution Research Centre, School of Biological, Earth and Environmental Sciences, University of New South Wales, Sydney, NSW, Australia &NewLine;  
^3^ School of Life and Environmental Sciences, University of Sydney, Sydney, NSW 2006, Australia  &NewLine;  
^4^ Department of Biology, Lancaster University, Liverpool, UK &NewLine;  
$\dagger$ contributed equally &NewLine;  
$\ddagger$ corresponding author, daniel.noble@anu.edu.au &NewLine;

@@

Climate change is expected to result in warmer and more variable thermal environments globally. Greater thermal variability is expected to result in strong selection pressures leading to genetic adaptation and/or the evolution of adaptive phenotypic plasticity. Such responses depend on genetic and phenotypic variability. However, most work has focused on changes in mean phenotypic responses to climate warming ignoring how temperature may also change phenotypic variability. Phenotypic variability may be particularly important at extreme, high temperatures, which would facilitate selection of resistant individuals or promote plasticity (acclimation) and thereby increase resilience to heat waves. Using newly developed effect size estimates and meta-analysis (>1900 effects from 226 species), we show that across habitats relative variance in physiological rates decreased at higher temperatures. Freshwater ectotherms are capable of acclimating and have the smallest reductions in relative variance. Marine organisms also showed a capacity to acclimate to higher temperatures, but capacity for plasticity traded-off with a reduction in relative variance in physiological rates at higher temperatures. Relative variance reductions were particularly pronounced for terrestrial ectotherms, and this coincided with a lack of capacity for acclimation, highlighting the vulnerability of terrestrial ectotherms to climate change. Neither life-history stage nor past climate explained effect variability. Our results show that beneficial acclimation responses may trade-off with reductions in physiological rate variance. This trade-off could constrain evolutionary responses to climate change and reduce the potential benefits of portfolio effects. These findings have important evolutionary and ecological ramifications that affect our understanding of how climate change will impacts populations now and in the future.
